The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-08-05 paper.qmd
finished trust and confidence section

## Why this approach may (or may not) work
The difficulty in incorporating uncertainty into a visualisation is frequently mentioned but seldom explained. For example @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty but did not explain why. Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006]. The problem with this approach to uncertainty visualisation is that it treats uncertainty the same as we would any other variable. However, @fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppresion. If we cannot treat uncertainty as any other variable, what should we treat it as? We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.

@@

### Trust and confidence

@moritz2017trust did some trust EDA stuff? check the paper

A similar measure to trust is using "confidence" in an extracted value or a decision. Interestingly, "confidence" is also used to try and capture the clarity of a message in a normal visualisation. Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation.

Trust is a by product of displaying uncertainty rather than the goal of it, and viewing the relationship in the converse direction can lead to misguided research. Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018]. @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy". These authors are optimising *trust* rather than *transparency*, which means they opt to leave out uncertainty information when it does exactly what it is supposed to, decrease certainty in conclusions.

Science communication should be primarily concerned with accuracy. Setting trust and risk-aversion as the variables of interest implicitly encourages statisticians to set trust and risk-aversion as the primary goals of communication. The issue of trust being divorced from trustworthiness has been commented on by other authors [@ONeill2018], however the issue still persists in the uncertainty visualisation literature. @Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust. They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive. The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy, but conflating trust and the use of a prediction implied uncertainty information should somehow influence participants to use their own prediction, even though a prediction being uncertainty does not necessarily mean it is incorrect. Despite this, the authors seemed to assume that the uncertainty information *should* have an influence on that, showing they had not deeply considered *how* uncertainty information should influence the choices of the participants. (*Cite Gap 18: examples of studies where authors measure trust*)

### Vague questions
There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations. These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs. This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].

@@

### Deterministic questions
Another method used by authors is to ask a deterministic question about a random event. @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions. In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution. Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage. Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. The flip side of this is asking participants for a deterministic answer to a probabilistic question. @Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference. Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B. The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.

>>>>

## Why this approach may (or may not) work
The difficulty in incorporating uncertainty into a visualisation is frequently mentioned but seldom explained. For example @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty but did not explain why. Many authors seem to believe uncertainty visualisation is a simple high-dimensional visualisation problem, as the difficulty comes from working out how to add uncertainty into already existing graphics [@Griethe2006]. The problem with this approach to uncertainty visualisation is that it treats uncertainty the same as we would any other variable. However, @fig-bivariate makes it clear that simply including uncertainty as a variable is insufficient to perform signal-suppresion. If we cannot treat uncertainty as any other variable, what should we treat it as? We need to understand what uncertainty actually *is*, in order to understand how to integrate it into a visualisation.

@@

### Trust and confidence
Trust is a by-product of displaying uncertainty and it commonly measured in uncertainty evaluation studies [@Hullman2019]. Considering trust, and not transparency, as the metric of importance in uncertainty communication can lead to a questionable subtext that argues against transparency, something that has been noticed by several other authors [@Spiegelhalter2017; ONeill2018]. Science communication should be primarily concerned with accuracy.

Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication. @Hullman2020a found that author simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy". This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions.

The disconnect between trust and transparency extends to evaluation experiments as well. @Zhao2023 displayed a several variations of a visualisation of a model prediction and its uncertainty and took participants using the model prediction as a sign of trust. They reported that visualising uncertainty information caused participants to trust the model in the low variance case, but the results in the high variance case were inconclusive. The discussion made it clear the authors thought the uncertainty information should make the visualisation more trustworthy. However, this perspective conflates transparency and trustworthiness. If a model has very high uncertainty because of a very low sample size, the heuristics a human might use to make the same decision could be more accurate.

Another metric that is similar to trust is the participants confidence in their decision or extracted value. Confidence has many of the same issues as trust, but it has an additional confounding factor. In non-uncertainty visualisation evaluation experiments, "confidence" is used as a proxy for the clarity of the visualisation. Confidence cannot simultaneously be a measure of clarity of visualisation *and* a way to capture the uncertainty expressed in a visualisation. Uncertainty visualisations conflate these two measures when they ask about confidence.

### Questions that attempts to capture signal-suppression
There is also a swath of studies that are aware a question that boils down to a value extraction experiment, or a question that should be answered by literally ignoring uncertainty information is not what we want when we consider uncertainty visualisations. These papers often try to ask a question that should utilise both the uncertainty and signal in the response, however this is rarely what actually occurs. This method typically results in in cryptic or confusing questions that create a large amount of noise on the interpretation side of the analysis [@Hullman2016].

@@

Another method used by authors is to ask a deterministic question about a random event. @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. This inclusion of determinism seems to cause the authors to stumble themselves, as they are not consistent with their assumptions. In their first experiment @Padilla2017 indicated the correct answer was to assume that the storm was equally intense no matter how far from the centre of the distribution an oil rig was, however answering their third experiment correctly hinged on assuming the intensity of the storm at a particular point (which in this experiment they phrased as damage) *does* change in intensity as you move away from the centre of the distribution. Given these conflicting assumptions, it is unclear how the participants were supposed to adjust the probabilistic path information to answer a deterministic question about which oil rig would receive the most damage. Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. The flip side of this is asking participants for a deterministic answer to a probabilistic question. @Correll2014 asked participants "how likely is candidate B to win the election?" when the two distributions indicated voter preference. Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1=Outcome will be most in favour of A to 7=Outcome will be most in favour of B. The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.
